Forecasting the 2024 U.S. Presidential Election: A Poll-Based Approach

### 2024-11-03 paper/paper.qmd
Updated Swing State Definition

---
title: "Forecasting the 2024 U.S. Presidential Election: A Poll-Based Approach"
subtitle: "Kamala Harris Poised for Victory"
author: 
  - Xinxiang Gao
  - Ariel Xing
  - John Zhang^[The authors are listed in alphabetical order by last name.]
thanks: "Code and data are available at: [https://github.com/xgao28/election_forecast](https://github.com/xgao28/election_forecast)."
date: today
date-format: long
abstract: "This paper presents a forecasting model for the 2024 U.S. Presidential Election, using polling data to analyze voter support for Kamala Harris and Donald Trump. Our approach combines a Baseline Model, which captures national polling trends, with a Primary Model focused on key swing states. Findings suggest a national advantage for Harris, with critical leads in swing states that could secure her a projected 292 electoral votes. This dual-model approach underscores the influence of state-level dynamics on national outcomes, offering insights into how polling data can more accurately capture voter behavior across a diverse electoral landscape."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(skimr)
library(knitr)
library(GGally)
library(dplyr)
library(broom)
```

@@

The U.S. Presidential Election is a critical event shaping the country’s political direction, often influenced by a complex interplay of public opinion, socio-economic factors, and electoral mechanisms. In recent election cycles, the predictive accuracy of polling has been a topic of significant discussion, as polls not only gauge public sentiment but also influence campaign strategies, media narratives, and voter perceptions. However, challenges such as non-response bias, methodological inconsistencies, and a misalignment between the popular vote and electoral outcomes highlight the need for a robust forecasting model that captures both national and state-specific trends.

This paper aims to address these challenges by developing a polling-based forecasting model for the 2024 U.S. Presidential Election, with a focus on the two main candidates, Kamala Harris and Donald Trump. Using a dual-model approach, we employ a **Baseline Model** to provide a broad national outlook and a **Primary Model** to examine state-level dynamics, especially in key swing states. The Baseline Model aggregates national polling data to predict which candidate would win a majority of popular support, while the Primary Model applies a more nuanced, state-level analysis aligned with the electoral college structure. This combination seeks to address the gap in existing models, which often fail to account for the specific impact of swing states on electoral outcomes, despite their pivotal role in determining the presidency.

Our analysis indicates that while Kamala Harris shows a lead in national polling, as reflected in the Baseline Model, the decisive factors lie within swing states such as Pennsylvania, Michigan, and Wisconsin, where the Primary Model highlights Harris’s slight advantages. The state-level predictions from the Primary Model suggest a possible electoral college win for Harris, projecting her to receive 292 electoral votes, above the 270-vote threshold needed to secure the presidency. This finding underscores the unique nature of U.S. elections, where winning the national popular vote does not necessarily translate to an electoral college victory. Our approach, by leveraging poll data with attention to poll quality, transparency, sample size, and geographic variation, seeks to provide a nuanced and accurate forecast that reflects both general sentiment and the regional intricacies of the election.

The structure of this paper is as follows: @sec-data provides an overview of the dataset, detailing the polling data sources, variables, and preprocessing steps. @sec-model outlines the modeling approach, presenting the rationale behind the Baseline and Primary Models and discussing their respective contributions to the analysis. @sec-results presents the results, including national and state-level predictions, followed by @sec-disc, which explores the implications of these findings, discusses limitations, and suggests future research directions. An appendix provides additional methodological details and diagnostics.

## Estimand

@@

@@

### Baseline Model (Winner-Take-All National Model)  
In our baseline model, we predict the election winner by using national polling data and assigning all electoral votes to the candidate with the higher projected percentage on election day. This "winner-take-all" approach assumes that the candidate leading in national polling would theoretically secure all electoral votes.

@@

### 2. Swing State-Based Model  
This model forecasts electoral outcomes by projecting each candidate’s polling results in individual swing states, then combining those results with electoral votes from states that are already solidly in favor of one candidate or the other.

@@

\appendix
# Appendix {-}

# Methodology Analysis of The Economist/YouGov Poll (October 6–7, 2024)

@@

# Methodology and Survey Design for 2024 U.S. Presidential Election Forecast

## Overview
This methodology presents an election forecasting plan with a budget of $100,000 focusing on swing states which are states where election outcomes are uncertain, making them crucial for forecasting the Electoral College result. Drawing on insights from YouGov and  Annual Review of Economics regarding survey methods, we chose an online format to improve accessibility and minimize selection bias, allowing respondents to participate at their convenience. This approach is especially suitable for students and working-age individuals, who may have limited availability during traditional hours. Mobile technology further enhances engagement, reaching populations that are otherwise challenging to access, such as seniors and mobile-first respondents. To maintain engagement and prevent drop-offs, we designed a concise survey with digital incentives to attract a broad participant base across income levels.
By incorporating techniques from both institutions, our methodology is designed to be cost-effective and demographically balanced, producing reliable predictions for the 2024 U.S. Presidential Election.

>>>>

---
title: "Forecasting the 2024 U.S. Presidential Election: A Poll-Based Approach"
subtitle: "Kamala Harris Poised for Victory"
author: 
  - Xinxiang Gao
  - Ariel Xing
  - John Zhang^[The authors are listed in alphabetical order by last name.]
thanks: "Code and data are available at: [https://github.com/xgao28/election_forecast](https://github.com/xgao28/election_forecast)."
date: today
date-format: long
abstract: "The upcoming U.S. Presidential Election on November 5th features a closely contested race between Vice President Kamala Harris and former President Donald Trump. This paper uses a poll-of-polls approach, combined with a dynamic linear regression model that updates daily with the latest data, to forecast the popular vote and outcomes in seven battleground states: Arizona, Georgia, Nevada, North Carolina, Michigan, Wisconsin, and Pennsylvania. Our analysis currently projects that Vice President Harris will win the popular vote, 48.19% to 46.01%, and secure the electoral college with 302 votes to Trump’s 236, by winning five of these seven swing states: Nevada, North Carolina, Michigan, Wisconsin, and Pennsylvania. By aggregating multiple polls rather than relying on a single poll, our model provides a more robust prediction, supporting the likelihood of a Harris victory."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(skimr)
library(knitr)
library(GGally)
library(dplyr)
library(broom)
library(styler)
```

@@

The race between Vice President Kamala Harris and former President Donald Trump for the 2024 presidential election will conclude on November 5, 2024. President Joe Biden's decision on July 21, 2024, to withdraw from this year’s race led to Vice President Kamala Harris becoming the official Democratic nominee, supported by Biden and confirmed through a virtual roll call on August 2. Even before Biden’s withdrawal, polling organizations had gauged public opinion on a potential matchup between Harris and former President Donald Trump. Early polls showed Harris trailing Trump, but her support increased following Biden's exit, making the outcome of this election uncertain.This paper employs a poll-of-polls approach and a dynamic linear model to project each candidate's likely support on election day, both nationally and within swing states.

@@

We apply a dynamic linear regression model named “ Baseline Model” to provide a national overview and another dynamic linear regression model named “Primary Model” to assess state-level dynamics, especially in key battlegrounds. The Baseline Model aggregates national polling data to estimate popular vote share, while the Primary Model offers a state-specific analysis aligned with the electoral college structure. This approach addresses a gap in existing models that often overlook the unique impact of swing states on electoral outcomes.
Our model projects that Vice President Harris will secure 48.19% of the popular vote, compared to 46.01% for Trump. However, as U.S. elections are decided by the electoral college rather than the popular vote, a candidate can win the presidency without a majority of popular support. There are seven swing states, Arizona, Georgia, Nevada, North Carolina, Michigan, Wisconsin, and Pennsylvania, holding a collective 93 electoral votes, are determining factors to the election’s outcome. Excluding swing states, Harris would receive 226 electoral votes, while Trump would receive 219. According to our Primary Model, Harris is expected to capture five of the seven swing states: Nevada, North Carolina, Michigan, Wisconsin, and Pennsylvania, yielding a total of 302 electoral votes and thus securing the presidency. Trump would win two swing states, Arizona and Georgia, ending with 236 electoral votes. Although Harris's lead in North Carolina is narrow (by just 0.036%), which may switch to voting for Trump in the future, but 19 votes in this state in the face of overwhelming vote margins won't have much of an impact on the outcome.

Predicting the next U.S. president holds significant implications for the nation’s future and its role on the global stage. The elected leader will influence domestic policy on economic growth, healthcare, and social issues, while also shaping international relations, defense strategies, and trade policies. In an era of geopolitical uncertainty, including conflicts and economic fluctuations, understanding potential election outcomes can help inform the public, policymakers, and global leaders, providing a foundation for navigating the challenges ahead.

The structure of this paper is as follows: @sec-data provides an overview of the dataset, detailing the polling data sources, variables, and preprocessing steps. @sec-model outlines the modeling approach, presenting the rationale behind the Baseline and Primary Models and discussing their respective contributions to the analysis. @sec-results presents the results, including national and state-level predictions, followed by @sec-disc, which explores the implications of these findings, discusses limitations, and suggests future research directions. The three appendices provide a detailed exploration of key aspects relevant to the 2024 U.S. Presidential Election. @sec-swingsolid explains the distinction between swing states and solid states, highlighting their importance in election outcomes. @sec-meth and @sec-survey delve into the methodology and design of the Economist/YouGov poll and an election forecasting survey, outlining sampling methods, data validation processes, and strategic considerations for accurate and reliable polling results.

@@

## Baseline Model (Winner-Take-All National Model)  
In our baseline model, we predict the election winner by using national polling data and assigning all electoral votes to the candidate with the higher projected percentage on election day. This "winner-take-all" approach assumes that the candidate leading in national polling would theoretically secure all electoral votes.

@@

## Swing State-Based Model  
This model forecasts electoral outcomes by projecting each candidate’s polling results in individual swing states, then combining those results with electoral votes from states that are already solidly in favor of one candidate or the other.

@@

\appendix
# Swing State and Solid State{#sec-swingsolid}
In the context of U.S. presidential elections, states are often classified as either "swing states" or "solid states" based on their voting behavior and predictability. Swing states, sometimes referred to as "battleground states," are those where neither major political party has a consistent or dominant advantage. This makes their electoral outcomes highly competitive and crucial for candidates aiming to secure a majority in the Electoral College. These states can shift between supporting Republican and Democratic candidates from one election to the next, depending on factors such as voter turnout, shifting demographics, and regional issues. Examples of prominent swing states for the 2024 election include Arizona, Georgia, Pennsylvania, Michigan, Nevada, North Carolina, and Wisconsin, as these states have shown significant variability in their support for both parties in recent election cycles (@citeSwing1) (@citeSwing2).

In contrast, solid states are those that consistently support one political party by a substantial margin, making their outcomes highly predictable in presidential elections. For instance, states like California and New York are considered solidly Democratic, while states like Alabama and Idaho are reliably Republican. These solid states play a lesser role in the competitive dynamics of presidential campaigns, as their electoral votes are typically seen as secure for the party that historically dominates them. Consequently, candidates often focus their campaign efforts, resources, and messaging on swing states where the race is closer, since these states ultimately have the potential to decide the outcome of the election (@citeSwing1).

# Methodology Analysis of The Economist/YouGov Poll (October 6–7, 2024) {#sec-meth}

@@

# Methodology and Survey Design for 2024 U.S. Presidential Election Forecast {#sec-survey}

## Overview
This methodology outlines an election forecasting plan with a budget of $100,000, focusing on swing states where election outcomes are uncertain, making them essential for forecasting the Electoral College result. Drawing on research from YouGov and the Annual Review of Economics on survey methods, we selected an online format to improve accessibility and reduce selection bias, allowing respondents to participate at their convenience(@citeStantcheva2023). This approach is particularly suited for students and working-age individuals who may have limited availability during traditional hours. Mobile technology also increases participation, reaching groups that are otherwise difficult to access, such as seniors and individuals who primarily use mobile devices. To encourage participation and reduce drop-offs, we created a concise survey with digital incentives to attract a broad range of respondents across income levels.
By incorporating methods from both sources, our approach is designed to be cost-effective and demographically balanced, offering reliable predictions for the 2024 U.S. Presidential Election.
